Investigation of the activities performed by experimental researchers in a software engineering lab using an ethnographic methodology

### 2023-12-24 tex/conclusion.tex
Some format and figure changes
Here there are some changes in the entire document.

We do not claim that our findings are commonplace in the empirical community. They may be particular to the experimental research group where we performed the ethnographic study. However, some clues evidence that peculiar behaviors are usual in the lab and the field. For instance, natural sciences researchers have complained about the replication problem \cite{hines2014sorting}. At least in the sciences (maybe not in SE), replication frequently fails due to experimental setup and conduction changes. These changes are motivated by the researchers' tacit knowledge \cite{Polanyi-1996-tacit-k}, \cite{Shull-2002-replicating-SE-experiments-tacit-k}, and consequently, protocols or experiment reports do not declare such changes. Experiment variations happen so frequently that some disciplines coined a (comic) term: \textquotedblleft lab mythology \textquotedblright~\cite{ruben2011experimental} \cite{loukides2015beyond}. Our study has shown that such tacit knowledge is present in SE and manifests relatively superficially, e.g., when researchers from different experimental families discuss their mental models.

>>>>

We do not claim that our findings are commonplace in the empirical community. They may be particular to the experimental research group where we performed the ethnographic study. However, some clues evidence that peculiar behaviors are usual in the lab and the field. For instance, natural sciences researchers have complained about the replication problem \cite{hines2014sorting}. At least in the sciences (maybe not in SE), replication frequently fails due to experimental setup and conduction changes. These changes are motivated by the researchers' tacit knowledge \cite{Polanyi-1996-tacit-k}, \cite{Shull-2002-replicating-SE-experiments-tacit-k}, and consequently, protocols or experiment reports do not declare such changes. Experiment variations happen so frequently that some disciplines coined a (comic) term: \textquotedblleft lab mythology \textquotedblright~\cite{ruben2011experimental} \cite{loukides2015beyond}. Our study has shown that such tacit knowledge is present in SE and manifests superficially, e.g., when researchers from different experimental families discuss their mental models.

### 2023-12-24 tex/research-execution.tex
Some format and figure changes
Here there are some changes in the entire document.

\section{Ethnographic study}\label{sec-reseach-execution}
We report the results of the ethnographic study chronologically: First, we describe the analysis of the RGUS's preferred literature sources and review the RGUS' specific experimental material. Later, we comment on our experiences replicating an existing experiment. Finally, we describe how we formalized the information into concept and process models using interviews, focus groups, and workshops.

\subsection{First Approximation}
In the first stages of the investigation, the researchers did not have an ethnographic methodology in mind but something less demanding, in line with existing qualitative studies, e.g., \cite{Johanssen-2019-Continuous-SE-support}, \cite{Strandberg-2019-Ethical-Interviews-SE}, and \cite{Yang-2021-interview-study-developers}. Accordingly, the researchers conducted interviews to clarify the RGUS' knowledge and convert it directly into models.

The interviews with RGUS members provided only public, limited knowledge about experimentation. The experimenters believed that interviews were unnecessary because the researchers could acquire knowledge by studying the literature that the experimenters considered the standard, e.g., \cite{Wohlin-2000-Experimentation-SE}. Hence, we did study the literature too.

However, communication became problematic when the interviews progressed into specialized areas, e.g., concrete families of experiments. For example, the experimenters supported their storytelling using specific experimental materials the researchers did not have. Nor did the researchers know to understand the experimenters' discussions. At this point, the need for a different research methodology --ethnography-- became apparent. Consequently, the researchers switched minds and continued at a slower pace. We carried out additional activities to understand the standard literature and experimental material. The interaction with the experimenters also changed. For example, the researchers participated first as experimental subjects and later as team members in one of the RGUS' families of experiments.

@@

The experimental materials evolve. Experiments vary widely in documentary support, independently of their date. The experimental family explains the differences more significantly than the time; the materials related to different families of experiments show significant differences, e.g., digital raw data versus physical papers written with raw data; or automatic tools to measure variables versus manual tools.

Concerning the experimental process, all experimental materials describe the activities carried out generally. The execution order and the relationship between experimental activities are unclear; they rely on the experimenters' tacit knowledge. Terminological diversity also affects experimental materials, especially among families of experiments.

@@

The existence of explicit experimental materials such as questionnaires, task descriptions, and programs significantly eased experiment conduction since the researchers defined in advance all relevant aspects. The measurement was the most complex and controversial aspect. Let us say that the experimental subjects return test cases, but these test cases are manually interpreted and not run. The experimenters' cooperation was again needed to complete the measurement. Even if the measurement procedure looked controversial, it made complete sense after discussion. We learned that the experimental protocols depend on the experimental task, which depends on the research area. The remaining steps, e.g., calculating measurements from raw data and formatting, did not imply any uncertainty. It was possible to reproduce the duplicate files used in the baseline experiment.

@@

However, we were not completely satisfied with the ethnographic study. We were able to replicate \textbf{one} experiment in \textbf{one} concrete SE area, but nothing guaranteed that we would not need the experimenters' help in a hypothetical subsequent replication \cite{Juristo-2012-replication-SE} \cite{Gomez-2014-understanding-replication}. The presence of tacit knowledge represented a severe threat to validity.

\subsubsection{Semi-structured interviews}
We decided to organize a new series of semistructured interviews with the experimenters. The key difference with the first approximation was that we would not discuss the SE experimental process \textit{in general}, but we would focus on replicating \textit{specific families of experiments}. Given the experience acquired during the replication, we believed this strategy could expose the tacit experimenters' knowledge.

@@

\begin{figure}[htbp!]
	\centering
	\includegraphics[width=\columnwidth]{images/Boseto-Proccess}
	\caption{example of a sketch made by experimenters.}
	\label{fig-proceso-exp-boseto}
\end{figure}

@@

\subsubsection{Preliminary models}\label{sec:preliminary-models}
We created process and concept models. We also created a workflow to represent the relations between activities that appeared during the investigation. Please notice that we introduce the outstanding characteristics of the models only because the ethnographic research did not end at this point yet. The models and the workflow have to be validated by the experimenters. We will provide full details in Section~\ref{sec-results}.

\begin{itemize}
	\item \textit{Preliminary} experimental process model: We realized that the RGUS conducts experiments that resemble the processes we use in SE, e.g., ISO/IEC 12207 \cite{ISO-IEC-IEEE-12207}. In the end, they all are processes. Hence, we abstracted six high-level experimental processes: support, generation of pieces of knowledge, publications, synthesis, basic, and organization. Each process has several related activities. For example, the \textit{basic process} includes the activities: pre-execution (context adaptation, experiment redesign, and experiment design), execution (pre-session, in-session, and post-session), and post-execution (raw data generation, data processing, data analysis, and data interpretation) (see this \href{https://zenodo.org/record/7102301#.Yyt0GOzMLUI}{\ul{link}}).

@@

As we said above, we organized the semistructured interviews around families of experiments. It was a fortunate mistake. We found out that the families of experiments are pretty homogeneous. The experimenters involved in the family quickly agree and provide a coherent picture of the experimental process. However, experimenters from different families disagree on many aspects, particularly the specific arrangements used in the experimental families. For instance, the experimenters never agreed on the activities related to the experimental design.

>>>>

\section{Ethnographic study}\label{sec-reseach-execution}
We report the results of the ethnographic study chronologically. First, we describe the analysis of the RGUS's preferred literature sources and review the RGUS' specific experimental material. Later, we comment on our experiences replicating an existing experiment. Finally, we describe how we formalized the information into concept and process models using interviews, focus groups, and workshops.

\subsection{First Approximation}
In the first stages of the investigation, the researchers did not have an ethnographic methodology in mind but something less demanding, in line with existing qualitative studies, e.g., \cite{Johanssen-2019-Continuous-SE-support}, \cite{Strandberg-2019-Ethical-Interviews-SE}, and \cite{Yang-2021-interview-study-developers}. Accordingly, the researchers conducted interviews to clarify the RGUS knowledge and convert it directly into models.

The interviews with RGUS members provided only limited public knowledge about experimentation. The experimenters believed that interviews were unnecessary because the researchers could acquire knowledge by studying the literature that the experimenters considered the standard, e.g., \cite{Wohlin-2000-Experimentation-SE}. Hence, we studied the literature, too.

However, communication became problematic when the interviews progressed into specialized areas, such as concrete families of experiments. For example, the experimenters supported their storytelling using specific experimental materials the researchers did not have. Nor did the researchers know to understand the experimenters' discussions. At this point, the need for a different research methodology --ethnography-- became apparent. Consequently, the researchers switched minds and continued at a slower pace. We carried out additional activities to understand the standard literature and experimental material. The interaction with the experimenters also changed. For example, the researchers participated first as experimental subjects and later as team members in one of the RGUS families of experiments.

@@

The experimental materials evolve. Experiments vary widely in documentary support, independently of their date. The experimental family explains the differences more significantly than the time; the materials related to different families of experiments show significant differences, e.g., digital raw data versus physical papers written with raw data or automatic tools to measure variables versus manual tools.

Concerning the experimental process, all experimental materials generally describe the activities carried out. The execution order and the relationship between experimental activities are unclear; they rely on the experimenters' tacit knowledge. Terminological diversity also affects experimental materials, especially among families of experiments.

@@

The existence of explicit experimental materials such as questionnaires, task descriptions, and programs significantly eased experiment conduction since the researchers defined all relevant aspects in advance. The measurement was the most complex and controversial aspect. Let us say that the experimental subjects return test cases, but these test cases are manually interpreted and not run. The experimenters' cooperation was again needed to complete the measurement. Even if the measurement procedure looked controversial, it made complete sense after discussion. We learned that the experimental protocols depend on the experimental task, which depends on the research area. The remaining steps, e.g., calculating measurements from raw data and formatting, did not imply any uncertainty. It was possible to reproduce the duplicate files used in the baseline experiment.

@@

However, we were not completely satisfied with the ethnographic study. We replicated \textbf{one} experiment in \textbf{one} concrete SE area. Still, nothing guaranteed that we would not need the experimenters' help in a hypothetical subsequent replication \cite{Juristo-2012-replication-SE} \cite{Gomez-2014-understanding-replication}. The presence of tacit knowledge represented a severe threat to validity.

\subsubsection{Semi-structured interviews}
We decided to organize a new series of semistructured interviews with the experimenters. The key difference with the first approximation was that we would not discuss the SE experimental process \textit{in general}. Still, we would focus on replicating \textit{specific families of experiments}. Given the experience acquired during the replication, we believed this strategy could expose the tacit experimenters' knowledge.

@@

\begin{figure}[htbp!]
	\centering
	\includegraphics[width=\columnwidth]{images/Process-Approximation-Draft}
	\caption{example of a sketch made by experimenters.}
	\label{fig-proceso-exp-boseto}
\end{figure}

@@

\subsubsection{Preliminary models}\label{sec:preliminary-models}
We created process and concept models. We also created a workflow to represent the relations between activities that appeared during the investigation. Please note that we introduce the outstanding characteristics of the models only because the ethnographic research has not ended yet. The models and the workflow have to be validated by the experimenters. We will provide full details in Section~\ref{sec-results}.

\begin{itemize}
	\item \textit{Preliminary} experimental process model: We realized that the RGUS conducts experiments that resemble the processes we use in SE, e.g., ISO/IEC 12207 \cite{ISO-IEC-IEEE-12207}. In the end, they are all processes. Hence, we abstracted six high-level experimental processes: support, generation of pieces of knowledge, publications, synthesis, basic, and organization. Each process has several related activities. For example, the \textit{basic process} includes the activities: pre-execution (context adaptation, experiment redesign, and experiment design), execution (pre-session, in-session, and post-session), and post-execution (raw data generation, data processing, data analysis, and data interpretation) (see this \href{https://zenodo.org/record/7102301#.Yyt0GOzMLUI}{\ul{link}}).

@@

As we said above, we organized the semistructured interviews around families of experiments. It was a fortunate mistake. The families of experiments are pretty homogeneous. The experimenters involved in the family quickly agree and provide a coherent picture of the experimental process. However, experimenters from different families disagree on many aspects, particularly the specific arrangements used in the experimental families. For instance, the experimenters never agreed on the activities related to the experimental design.

### 2023-12-24 tex/research-method.tex
Some format and figure changes
Here there are some changes in the entire document.

Ethnography identifies details that are ignored by or in other methods. The disadvantage is the time required for learning \cite{Sharp-2016-Ethnographic-Studies-ESE}. Thus, we opted for the study of a single experimental research group over a long period instead of working halfway with several experimental research groups over shorter periods. This decision was also motivated by the challenge of finding several open-minded experimental research groups willing to participate in this research. We selected a representative i.e., respected and well-known, research group to guarantee (to the extent possible) the reliability of the results, their validity, and the findings' scientific significance \cite{sjoberg-2007-future-empical-methods}.

@@

To avoid confusion, from this point on we will refer to the "ESE researchers" with whom we interact during the investigation as "the experimenters". We will talk about ourselves the authors as "the researchers". We authors/researchers play two different roles in the investigation:

\begin{itemize}
\item \textbf{Efra\'in R. Fonseca C.} and Edison G. Espinosa conducted the ethnographic study as part of their Ph.D. Edison G. Espinosa abandoned the ethnographic study early (during the investigation reported in Section~\ref{replication}) and changed the orientation of this Ph.D. This is the reason he is not authoring this paper. The personal opinions poured in Section~\ref{sec-reseach-execution} belong to these two researchers, particularly Efra\'in R. Fonseca C.
\item \textbf{Oscar Dieste} and \textbf{Natalia Juristo} supervised the investigation. Their views and opinions are not reflected in this manuscript.
\end{itemize}

\subsection{Context selection}
We have considered an experimental research group as "representative", i.e., a reliable source of information, when it fulfills the following selection criteria:

\begin{itemize}
	\item The experimental research group has extensive experience in ESE research.
	\item The ESE community recognizes the experimenters as specialists in their areas.
	\item The group has a large number of publications in journals of repute and internal documentation, i.e., the group has a verifiable track record.
\item One of the main drivers of this research is to explore the connection between replication and lab practices. We positively value that the experimental research group conducts replications or families of experiments.
\end{itemize}

@@

Ethnography information gathering is characterized by being cyclical, iterative, and incremental. The data acquisition process was based on a taxonomy of information gathering techniques, categorized according to the level of contact with the primary source of information. Such a taxonomy proposes three levels of information gathering: direct participation with the source (level 1), indirect participation with the source (level 2), and study of the experiment material without the participation of the source (level 3) \cite{Lethbridge-2005-data-collection-techniques-SE}. Table \ref{tbl-tecnica-fuente} describes the information gathering techniques and instruments and the corresponding information sources to which they can be applied.

@@

The order in which researchers contact information sources is essentially random. The frequency of contact depends on the researchers' need to answer research questions during the ethnographic research and the availability of each source.

@@

Ethnographic practice involves storing the raw data and then interpreting the findings \cite{Cooper-2007-Sharing-Data-Ethnographic-Research}. In this research, the data analysis procedure consists of four activities:

\begin{itemize}
\item \textbf{Preparation of materials}: The information sources are classified and categorized, so experimenters and researchers can manage them.
\item \textbf{Information acquisition}: While analyzing the information sources, researchers gradually acquire basic knowledge about experimentation, which helps them understand the experimenters' knowledge and behavior more easily.
\item \textbf{Verification of information}: This is probably the most complex activity. The knowledge obtained from the sources complements each other. We will use triangulation to explain the findings and identify new areas of inquiry. All findings are discussed with the experimenters to avoid misunderstandings and verify that the resulting activities are those that they carry out in practice.
\item \textbf{Generation of results}: Finally, we obtained the results as final products after several iterations of comparison, verification, and consensus between researchers and experimenters. The expected results will be models (conceptual and process) and practices derived from the interviews with the experimenters.
\end{itemize}

\subsection{Research documentation}\label{sub-sec-research-documentaition}
The raw data obtained in this study is organized as shown in Figure \ref{fig-info-study}. The tree represents a repository of around 20GB of information. The repository contains pictures, recordings, models, reports, and transcriptions, most in Spanish (the operational language used during the research). Spanish is widespread enough to utilize the repository with little effort. The final products, e.g., models, have been translated into English to improve the reader's understanding. The repository is not public because it contains personal data. However, the corresponding author will share the repository with interested researchers upon request.

>>>>

Ethnography identifies details that are ignored by or in other methods. The disadvantage is the time required for learning \cite{Sharp-2016-Ethnographic-Studies-ESE}. Thus, we opted to study a single experimental research group over a long period instead of working halfway with several experimental research groups over shorter periods. This decision was also motivated by the challenge of finding several open-minded experimental research groups willing to participate in this research. We selected a representative, i.e., respected and well-known, research group to guarantee (to the extent possible) the reliability of the results, their validity, and the findings' scientific significance \cite{sjoberg-2007-future-empical-methods}.

@@

To avoid confusion, from this point on, we will refer to the "ESE researchers" with whom we interact during the investigation as "the experimenters". We will talk about ourselves, the authors, as "the researchers". We, authors/researchers, play two different roles in the investigation:

\begin{itemize}
\item \textbf{Efra\'in R. Fonseca C.} and Edison G. Espinosa conducted the ethnographic study as part of their Ph.D. Edison G. Espinosa abandoned the ethnographic study early (during the investigation reported in Section~\ref{replication}) and changed the orientation of his Ph.D. This is the reason he is not authoring this paper. The personal opinions poured in Section~\ref{sec-reseach-execution} belong to these two researchers, particularly Efra\'in R. Fonseca C.
\item \textbf{Oscar Dieste} and \textbf{Natalia Juristo} supervised the investigation. Their views and opinions are not reflected in this manuscript.
\end{itemize}

\subsection{Context selection}
We have considered an experimental research group as "representative," i.e., a reliable source of information when it fulfills the following selection criteria:

\begin{itemize}
	\item The experimental research group has extensive experience in ESE research.
	\item The ESE community recognizes the experimenters as specialists in their areas.
	\item The group has many publications in journals of repute and internal documentation, i.e., the group has a verifiable track record.
\item One of the main drivers of this research is to explore the connection between replication and lab practices. We positively value that the experimental research group conducts replications or families of experiments.
\end{itemize}

@@

Ethnography information gathering is characterized by being cyclical, iterative, and incremental. The data acquisition process was based on a taxonomy of information-gathering techniques, categorized according to the level of contact with the primary source of information. Such a taxonomy proposes three levels of information gathering: direct participation with the source (level 1), indirect participation with the source (level 2), and study of the experiment material without the participation of the source (level 3) \cite{Lethbridge-2005-data-collection-techniques-SE}. Table \ref{tbl-tecnica-fuente} describes the information-gathering techniques and instruments and the corresponding information sources to which they can be applied.

@@

The order in which researchers contact information sources is random. The frequency of contact depends on the researchers' need to answer research questions during the ethnographic research and the availability of each source.

@@

The ethnographic practice involves storing the raw data and then interpreting the findings \cite{Cooper-2007-Sharing-Data-Ethnographic-Research}. In this research, the data analysis procedure consists of four activities:

\begin{itemize}
\item \textbf{Preparation of materials}: The information sources are classified and categorized so experimenters and researchers can manage them.
\item \textbf{Information acquisition}: While analyzing the information sources, researchers gradually acquire basic knowledge about experimentation, which helps them understand the experimenters' knowledge and behavior more easily.
\item \textbf{Verification of information}: This is probably the most complex activity. The knowledge obtained from the sources complements each other. We will use triangulation to explain the findings and identify new areas of inquiry. All findings are discussed with the experimenters to avoid misunderstandings and verify that the resulting activities are those that they carry out in practice.
\item \textbf{Generation of results}: Finally, we obtained the results as final products after several iterations of comparison, verification, and consensus between researchers and experimenters. The expected results will be models (conceptual and process) and practices derived from the interviews with the experimenters.
\end{itemize}

\subsection{Research documentation}\label{sub-sec-research-documentaition}
The raw data obtained in this study is organized as shown in Figure \ref{fig-info-study}. The tree represents a repository of around 20GB of information. The repository contains pictures, recordings, models, reports, and transcriptions, most in Spanish (the operational language used during the research). Spanish is widespread enough to utilize the repository with little effort. The final products, e.g., models, have been translated into English to improve the reader's understanding. The repository is not public because it contains personal data. However, upon request, the corresponding author will share the repository with interested researchers.
